\section{Fixed-wing dynamics}
This first section will give the reader a quick overview of some common definitions and terminology. 
A general purpose 6-DOF model of a fixed-wing \ac{uav} will also be derived. 
\subsection{General definitions and terminology}
We begin with establishing some common definitions and
 terminology which will be used throughout this thesis. These definitions are 
 used in many other works related to fixed-wing aircraft, such as \cite{uav_dynamics_wind}, \cite{uav_dynamics_modeling}, 
\cite{fighter_aircraft_mpc}. 

\subsubsection{Coordinate reference frames}
Three different coordinate reference frames are used, an inertial frame, a body frame and a wind reference frame.
This is convenient since some sensors, like \ac{gps} report their values in the inertial frame
 while others, like \ac{imu} sensors report values in the body frame. Other advantages of the different references
will be made apparent when formulating the dynamic equations.

\begin{definition}[Inertial frame]
    The inertial frame, denoted with subscript $I$ is fixed relative to the earth.
    A position vector in the inertial frame is defined in the \ac{ned} order as
    \begin{equation}
        \vec{p}_I = (x_N, y_E, -z_H)
    \end{equation}
    where $x_N$ points in the north direction, $y_E$ points east and $z_H$ points down towards the earth,
    in order to form a right hand positive coordinate system.
\end{definition}

\begin{definition}[Body frame]
    The body frame, denoted with subscript $B$ is fixed in the \ac{uav} center of gravity.
    A position vector in the body frame is defined as
    \begin{equation}
        \vec{p}_B = (x, y, z)
    \end{equation}
    where $x$ points forward through the \ac{uav}, $y$ points to the right and $z$ points down as shown in Figure TODO
\end{definition}

\begin{definition}[Wind reference frame]
    The wind reference frame, denoted with subscript $\windspd$ is related to the current direction of motion
    through the air.
    A position vector in the wind reference frame is defined as
    \begin{equation}
        \vec{p}_W = (x_w, y_w, z_w)
    \end{equation}
    where $x_w$ points in the same direction as the current velocity vector $\vel_I$, 
    $y_w$ points to the right of $x_w$ and $z$ points down relative $x_w$ and $y_w$.
\end{definition}

\subsubsection{Attitude representation}
The attitude of the \ac{uav} is represented by the \textit{Euler angles}. 

\begin{definition}[Euler angles]
The Euler angle vector is defined as
\begin{equation}
    \vec{\Phi}=(\phi, \theta, \psi)
\end{equation}
where the \textit{roll angle} $\phi$ is rotation around the north inertial axis, 
the \textit{pitch angle} $\theta$ is rotation around the east inertial axis and
the \textit{yaw angle} $\psi$ is rotation around the downwards inertial axis.
\end{definition}

The relationship between coordinates in the body frame and inertial frame is given
 by the rotation matrix

\begin{equation}\label{eq:r_i_b}
\mathcal{R}^I_B = \mathcal{R}^x_\phi\mathcal{R}^y_\theta\mathcal{R}^z_\psi\\
=
\begin{bmatrix}
    1 & 0 & 0 \\
    0 & \cos\phi & \sin\phi \\
    0 & -\sin\phi & \cos\phi
\end{bmatrix}
\begin{bmatrix}
    \cos\theta & 0 & -\sin\theta \\
    0 & 1 & 0 \\
    \sin\theta & 0 & \cos\theta
\end{bmatrix}      
\begin{bmatrix}
    \cos\psi & \sin\psi & 0 \\
    -\sin\psi & \cos\psi & 0 \\
    0 & 0 & 1
\end{bmatrix}
\end{equation}  
This attitude representation is not defined for $\theta=\pm\pi/2$. However, such attitudes
 were deemed very unlikely in this work as the main focus is on level flight scenarios.

\subsubsection{Aerodynamic angles}
The \textit{aerodynamic angles} relate the wind reference frame to the other coordinate frames.

\begin{definition}[Aerodynamic angles]
We first define the body frame velocity vector as 
\begin{equation}
    \vel_B=(u, v, w)
\end{equation}
The \textit{angle of attack} $\alpha$ and \textit{side slip} $\beta$ are then defined as
\begin{equation}
    \alpha=\arctan\frac{w}{u}
\end{equation}
\begin{equation}
    \beta=\arcsin\frac{v}{\airspd}
\end{equation}
where $\airspd=\sqrt{u^2+v^2+w^2}$
\end{definition}


\subsection{Dynamic equations}
In the following section the dynamic equations of a general fixed-wing aircraft will be derived. The 
equations are based on the following common state representation:
\begin{equation}
    \vec{x}=
    \begin{bmatrix}
        \vec{p}_I & \vel_B & \vec{\Phi} & \vec{\omega}
    \end{bmatrix}
\end{equation}
where $\vec{\omega}=(p, q, r)$ are the angular rates in the body reference frame. These can be divided
into translational and rotational dynamics which are presented separately.
\subsubsection{Translational dynamics}
The dynamic equation for the position $\vec{p}_I$ is directly given as
\begin{equation}\label{eq:dynamic_first}
    \dot{\vec{p}}_I=\mathcal{R}^I_B\vel_B
\end{equation}
Furthermore, Newtons second law of motion gives 
\begin{equation}\label{eq:uav_force}
    \vec{F}_{tot}=m\dot{\vel}_B + \vec{\omega}\times m\vel_B 
\end{equation}
where $m$ is the \ac{uav} mass and $\vec{F}_{tot}$ is the sum of all forces working on the \ac{uav}.
These can be divided into
\begin{equation}
    \vec{F}_{tot}=\vec{F}_g+\vec{F}_{aero}+\vec{F}_{thr}
\end{equation}
where $\vec{F}_g$ is the gravitational force, $\vec{F}_{aero}$ is the aerodynamic force and
$\vec{F}_{thr}$ is the thrust force. The thrust force is assumed to only work in the same direction
as the x-axis in the body frame, so that
\begin{equation}
    \vec{F}_{thr, B}=(T, 0, 0)
\end{equation}
\subsubsection{Rotational dynamics}
The dynamic equations for the Euler vector $\vec{\Phi}$ are given by
\begin{equation}
    \begin{bmatrix}
        \dot{\phi} \\
        \dot{\theta} \\
        \dot{\psi}
    \end{bmatrix}
    =
    \begin{bmatrix}
        1 & -\sin\phi\tan\theta & \cos\phi\tan\theta \\
        0 & \cos\phi & \sin\phi \\
        0 & -\sin\phi\sec\theta & \cos\phi\sec\theta
    \end{bmatrix}
    \begin{bmatrix}
        p \\
        q \\
        r
    \end{bmatrix}
\end{equation}
By again using Newtons second law for the moment we derive
\begin{equation}
    \vec{M}=\vec{I}\dot{\vec{\omega}} + \vec{\omega}\times\vec{I}\vec{\omega}
\end{equation}
The moment $\vec{M}$ is assumed to mainly come from aerodynamics.
\subsubsection{Aerodynamics}
The aerodynamic forces and moments generated require some further analysis. The aerodynamic
force $\vec{F}_{aero}$ can be written in the wind reference frame as
\begin{equation}
    \vec{F}_{aero}=
    \begin{bmatrix}
        -D \\
        Y \\
        -L
    \end{bmatrix}=
    \begin{bmatrix}
        \bar{q}SC_D \\
        \bar{q}SC_Y \\
        \bar{q}SC_L
    \end{bmatrix}
\end{equation}
where $\bar{q}=\frac{1}{2}\rho(h)V^2$ is the free-stream dynamic pressure, $S$ is the 
wing surface area and $C_D$, $C_Y$ and $C_L$ are dimensionless constants.
$D$, $Y$ and $L$ are called the \textit{drag}, \textit{sideforce} and \textit{lift} respectively. 
The corresponding coefficients are dependent on a number of variables, such as the aerodynamic angles
and airframe dependent parameters \cite{fighter_aircraft_mpc}.\\

The aerodynamic moments are usually defined in the body reference frame as
\begin{equation}\label{eq:dynamic_last}
    \vec{M}_{aero}=
    \begin{bmatrix}
        M_x \\
        M_y \\
        M_z
    \end{bmatrix}=
    \begin{bmatrix}
        \bar{q}SbC_l \\
        \bar{q}SbC_m \\
        \bar{q}SbC_n
    \end{bmatrix}
\end{equation}
where $b$ is the wing-span of the aircraft and $C_l$, $C_m$ and $C_n$ are dimensionless 
constants mainly dependent on the aerodynamic angles \cite{uav_dynamics_wind}. \\

\subsubsection{Wind effects}
The dynamic equations derived in \eqref{eq:dynamic_first}-\eqref{eq:dynamic_last} are under
the assumption that the air through which the \ac{uav} is travelling is at rest relative to the earth.
During real flights, wind will cause the air to move which has non-negligible effects on the \ac{uav}.
The wind vector is often defined in the inertial frame as
\begin{equation}
    \windvec=(w_N, w_E, w_H)
\end{equation}
$\windvec$ can be decomposed to
\begin{equation}
    \windvec=\windvec_{m}+\windvec_s
\end{equation}
where $\windvec_m$ is the mean wind and $\windvec_s$ is some stochastic process \cite{spline_trajectory}. In this work,
we only consider the mean wind.
The mean wind magnitude $\windspd=|\windvec|$ varies with altitude as
\begin{equation}
    \windspd(h)=W_{10}(\frac{h}{h_{10}})^a
\end{equation}
where $W_{10}$ is the wind measured at $h=10$ meters and $a$ is the Hellman exponent, which depends on
the shape and coastal location of the underlying terrain \cite{uav_dynamics_wind}. The wind can be 
incorporated in the dynamic equations of motion by extending \eqref{eq:dynamic_first} to
\begin{equation}
    \dot{\vec{p}}=\mathcal{R}^I_B\vel_B+\windvec
\end{equation}
Also, we need to extend \eqref{eq:uav_force} by replacing $\vel_B$ with
\begin{equation}
    \vel_g = \vel_B + \windvec_B
\end{equation} 
where $\vel_g$ is the body-fixed velocity relative the ground and $\windvec_B$ is the
wind vector in the body frame.

\subsubsection{Complete dynamics model}
The complete dynamic model is given by combining the above equations. It will not be 
written out here, but is given in \eg \cite{uav_dynamics_wind}. The dynamics are clearly nonlinear and quite complex. Extensive modeling and experimentation, such as described in \cite{aerodynamic_parameters}
is needed to identify the different parameters which describe the behaviour of a specific airframe.

It is often interesting to study the \textit{longitudal} and \textit{lateral} dynamics
separately, where \textit{longitudal} denotes movement in the vertical plane and
\textit{lateral} denotes movement out of the vertical plane. The decoupled state
vectors are given by
\begin{equation}
    \vec{x}_{lng}=(x_N, z_H, u, w, \theta, q)
\end{equation}
and
\begin{equation}
    \vec{x}_{lat}=(y_E, v, \phi, \psi, p, r)
\end{equation}
By introducing the polar inertial components $(V, \gamma)$ and writing the dynamic equations
in the wind reference frame the longitudal dynamics become
\begin{align}
    \dot{V}&=\frac{1}{m}[T\cos\alpha-D-mg\sin\gamma] \\
    \dot{\gamma}&=\frac{1}{mV}[T\sin\alpha + L - mg\cos\gamma] \\
    \dot{q}&=\frac{M_y}{I_{yy}} \\
    \dot{\alpha}&=\dot{\theta}-\dot{\gamma}=q-\frac{1}{mV}[T\sin\alpha + L - mg\cos\gamma]
\end{align}
Assuming level flight in the longitudal direction, \ie\  $u$, $V$ and $z_H$ are constant, the lateral dynamics can be formulated as
\begin{align}
    \dot{y}_E &= u\sin\psi + v\cos\phi\cos\psi \\
    \dot{\psi} &= r\cos\phi \\
    \dot{\phi} &= p \\
    \dot{v} &= \frac{Y_B}{m} + g\sin\phi - ru \\
    \dot{p} &= \frac{I_{zz}M_x + I_{xz}M_z}{I_{xx}I_{zz} - I^2_{xz}} \\
    \dot{r} &= \frac{I_{xz}M_x + I_{xz}M_z}{I_{xx}I_{zz} - I^2_{xz}} \\
\end{align}